Uniformly Ergodic Data-Augmented MCMC for Fitting Stochastic Epidemic Models to Incidence Data

### 2021-11-02 Manuscripts/Article/Article.tex

	It is straightforward to show that the Markov chain $\{(\theta^{(m)}, \z^{(m)})\}_m$ is harris ergodic and converges to the target posterior distribution $\pi(\theta, \z|\Y)$. For any $\pi$-integrable function $g$, the estimator
	$$\bar{g}_m := \frac{1}{m} \sum_{i=0}^{m-1} g(\theta^{(m)}, \z^{(m)})$$
	is therefore consistent for $E_\pi g$, regardless of the starting values of the chain. A Markov chain is uniformly ergodic if for some finite $M$ and positive constant $r<1$ the $n$-step transition kernel $P^n$ satisfies
	$$
	\Vert P^n(x,.)-\pi(.)\Vert_{TV} \le M r^n, \forall x \in \chi
	$$
	where $\Vert \mu \Vert_{TV}$ denotes the total variation norm of the measure $\mu$.
	Uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$ (ref), that is,
	$$\sqrt{m}(\bar{g}_m - E_\pi g) \Rightarrow N(0, \sigma^2)$$
	for some positive, finite $\sigma^2$. Furthermore, uniform ergodicity implies that the batch means and spectral variance estimators of $\sigma^2$ are consistent (ref).

@@

	Figure \ref{fig:E3_runtime} and \ref{fig:E3_accept} present the run time of the algorithm and the acceptance rate in the Metropolis-Hastings step. We observe that the run time increases with the population size and with $\rho$ and that, as expected, the acceptance rate decreases as the number of trajectories updated per iteration $\rho(S(0)+I(0))$ increases.
	Finally, Figure \ref{fig:E3_facet_ESSsecR0} shows the effect of $\rho$ on the auto correlation function for the parameter $\beta$ for the medium-size population process ($S(0) = 1,000$) for three values of $\rho$. We observe that updating half of the augmented data per iteration yields the lowest auto-correlation function across the values of $\rho$ considered. This suggests that, in this setting, updating the entire latent space in the Metropolis-Hastings step results in too many rejections, while updating a tenth or less prevents the chain from efficiently exploring the latent space.

	\begin{figure}
		\centering
		\begin{subfigure}[b]{0.42\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E3_runtime}
			\caption{Run time in minutes}
			\label{fig:E3_runtime}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E3_accept}
			\caption{Acceptance rate}
			\label{fig:E3_accept}
		\end{subfigure}
		\\
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E3_facet_ESSsecR0}
			\caption{ESS/sec for $R_0$
			%Acceptance rate of the proposed latent spaces in the Metropolis-Hastings step of the MCMC algorithm across different population sizes and values for $\rho$.
			}
			\label{fig:E3_facet_ESSsecR0}
		\end{subfigure}
		\caption{Impact of $\rho$ on the performance of the DA-MCMC across different population sizes.}
		\label{fig:E3}
	\end{figure}

>>>>

	It is straightforward to show that the Markov chain $\{(\theta^{(m)}, \z^{(m)})\}_m$ is harris ergodic and converges to the target posterior distribution $\pi(\theta, \z|\Y)$. For any $\pi$-integrable function $g$, the estimator
	$$\bar{g}_m := \frac{1}{m} \sum_{i=0}^{m-1} g(\theta^{(m)}, \z^{(m)})$$
	is therefore consistent for $E_\pi g$, regardless of the starting values of the chain. A Markov chain is uniformly ergodic if for some finite $M$ and positive constant $r<1$ the $n$-step transition kernel $P^n$ satisfies
	$$
	\Vert P^n(x,.)-\pi(.)\Vert_{TV} \le M r^n, \forall x \in \chi
	$$
	where $\Vert \mu \Vert_{TV}$ denotes the total variation norm of the measure $\mu$.
	Uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$ (Jones, 2004), that is,
	$$\sqrt{m}(\bar{g}_m - E_\pi g) \Rightarrow N(0, \sigma^2_g)$$
	for any initial distribution for some positive, finite $\sigma^2_g$. Furthermore, uniform ergodicity implies that the batch means and spectral variance estimators of $\sigma^2_g$ are consistent (ref).

@@

	Figure \ref{fig:E3_runtime} and \ref{fig:E3_accept} present the run time of the algorithm and the acceptance rate in the Metropolis-Hastings step. We observe that the run time increases with the population size and with $\rho$ and that, as expected, the acceptance rate decreases as the number of trajectories updated per iteration $\rho(S(0)+I(0))$ increases.
	Finally, Figure \ref{fig:E3_facet_ESSsecR0} shows the effect of $\rho$ on the mixing properties of the Markov chain. We observe for a population of $250$ individuals, updating the entire latent data in the MH step provides the best mixing; as the population size increases, the optimal value of $\rho$ decreases. This suggests that for a population of $1000$ individuals, updating the whole latent data results in too many rejections, while updating less than $10$\% makes the updates too small to efficiently explore the latent space.

	\begin{figure}
		\centering
		\begin{subfigure}[b]{0.42\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E3_runtime}
			\caption{Run time in minutes}
			\label{fig:E3_runtime}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E3_accept}
			\caption{Acceptance rate}
			\label{fig:E3_accept}
		\end{subfigure}
		\\
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E3_facet_ESSsecR0}
			\caption{ESS/sec for $R_0$
			%Acceptance rate of the proposed latent spaces in the Metropolis-Hastings step of the MCMC algorithm across different population sizes and values for $\rho$.
			}
			\label{fig:E3_facet_ESSsecR0}
		\end{subfigure}
		\caption{Impact of $\rho$ on the performance of the DA-MCMC in a population of size $250$ (solid), $500$ (dotted) and $1000$ (dashed).}
		\label{fig:E3}
	\end{figure}
